---
title: |
  \vspace*{\fill}
  \fontsize{36}{40}\selectfont \textbf {From Lab Tests to Quality Scores: A Neural Network Model for Red Wine Quality at GNAWC}
author: |
  \fontsize{24}{28}\selectfont Seif H. Kungulio
date: |
  \fontsize{18}{22}\selectfont January 17, 2026
  \ \vspace*{\fill}
output: 
  pdf_document:
    latex_engine: xelatex
    toc: false
    highlight: espresso
    include:
      in_header: resources/header.tex
header-includes:
  # Font and color setup
  - \usepackage{xcolor}
  - \definecolor{brand}{HTML}{990012}
  - \definecolor{subbrand}{HTML}{212121}
  - \usepackage{titlesec}
  - \titleformat{\section}{\Large\bfseries\color{brand}}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries\color{subbrand}}{\thesubsection}{0.75em}{}
  - \usepackage[most]{tcolorbox}
  - \tcbset{colback=white, colframe=brand, coltitle=white, fonttitle=\bfseries}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{\textcolor{brand}{From Lab Tests to Quality Scores}}
  - \fancyhead[R]{\textcolor{subbrand}{Seif H. Kungulio}}
  - \fancyfoot[C]{\thepage}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=8)

# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

#Install the following libraries if required
# List of required packages
required_pkgs <- c("tidyverse", "knitr", "janitor", "skimr", "glue",
                   "keras", "rsample", "recipes", "yardstick") 

# Install and load each package
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Data path
data_path <- "https://archive.ics.uci.edu/ml/machine-learning-databases/wine-quality/winequality-red.csv"

theme_set(theme_test(base_size = 14))
options(dplyr.summarise.inform = FALSE)
```

\pagenumbering{gobble}

\newpage
\thispagestyle{plain}
&nbsp;

\newpage
\pagenumbering{roman}
\tableofcontents

\newpage
\pagenumbering{arabic}

# **Business Understanding**
## Background
Great North American Wine Company (GNAWC) collects physicochemical lab measurements (e.g., acidity, sugar, alcohol, sulphates, pH) during production. Final quality ratings, however, are primarily determined by sensory evaluation, which can be subjective and costly.

## Problem Statement
Build a neural network model that predicts wine quality scores using physicochemical tests only, enabling earlier, standardized, data-driven screening before bottling/distribution.

## Objectives
* Predict quality scores accurately from lab measurements
* Detect potential low-quality batches earlier
* Reduce routine reliance on tasting panels for screening
* Improve quality decision consistency

## Success Criteria
* **Primary:** Low prediction error (MAE/RMSE) on a held-out test set
* **Operational:** Strong detection of low-quality wines (e.g., quality ≤ 5) via recall/ROC-AUC (optional classification view)
* **Business:** Practical thresholding rules for early screening (flagging likely low-quality batches)

\newpage
# **Data Understanding**
## Load Data
```{r load-data}
wine_data <- read_csv(data_path, col_names = TRUE, show_col_types = FALSE)
glimpse(wine_data)
head(wine_data, 10)
```

## Target Variable Check
```{r target-variable}
stopifnot("quality" %in% names(wine_data))

summary(wine_data$quality)
kable(wine_data$quality)

```

##