---
title: |
  \vspace*{\fill}
  \fontsize{36}{40}\selectfont \textbf {From Lab Tests to Quality Scores: A Neural Network Model for Red Wine Quality at GNAWC}
author: |
  \fontsize{24}{28}\selectfont Seif H. Kungulio
date: |
  \fontsize{18}{22}\selectfont January 17, 2026
  \ \vspace*{\fill}
output: 
  pdf_document:
    latex_engine: xelatex
    toc: false
    highlight: espresso
    include:
      in_header: resources/header.tex
header-includes:
  # Font and color setup
  - \usepackage{xcolor}
  - \definecolor{brand}{HTML}{990012}
  - \definecolor{subbrand}{HTML}{212121}
  - \usepackage{titlesec}
  - \titleformat{\section}{\Large\bfseries\color{brand}}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries\color{subbrand}}{\thesubsection}{0.75em}{}
  - \usepackage[most]{tcolorbox}
  - \tcbset{colback=white, colframe=brand, coltitle=white, fonttitle=\bfseries}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[L]{\textcolor{brand}{A Neural Network Model for Red Wine Quality at GNAWC}}
  - \fancyhead[R]{\textcolor{subbrand}{Seif H. Kungulio}}
  - \fancyfoot[C]{\thepage}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=12, fig.height=8)

# Set default CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

#Install the following libraries if required
# List of required packages
required_pkgs <- c("tidyverse", "knitr", "janitor", "skimr", "glue", "neuralnet", 
                   "rsample", "recipes", "yardstick", "GGally") 

# Install and load each package
for (pkg in required_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Set theme for ggplot2
theme_set(theme_test(base_size = 14))
options(dplyr.summarise.inform = FALSE)
```

\pagenumbering{gobble}

\newpage
\thispagestyle{plain}
&nbsp;

\newpage
\pagenumbering{roman}
\tableofcontents

\newpage
\pagenumbering{arabic}

# **Business Understanding**
## Background
Great North American Wine Company (GNAWC) collects physicochemical lab measurements (e.g., acidity, sugar, alcohol, sulphates, pH) during production. Final quality ratings, however, are primarily determined by sensory evaluation, which can be subjective and costly.

## Problem Statement
Build a neural network model that predicts wine quality scores using physicochemical tests only, enabling earlier, standardized, data-driven screening before bottling/distribution.

## Objectives
* Predict quality scores accurately from lab measurements
* Detect potential low-quality batches earlier
* Reduce routine reliance on tasting panels for screening
* Improve quality decision consistency

## Success Criteria
* **Primary:** Low prediction error (MAE/RMSE) on a held-out test set
* **Operational:** Strong detection of low-quality wines (e.g., quality ≤ 5) via recall/ROC-AUC (optional classification view)
* **Business:** Practical thresholding rules for early screening (flagging likely low-quality batches)

\newpage
# **Data Understanding**
## Load Data
```{r load-data}
wine.df <- read.csv("resources/wine.csv")
```

## Data Dimensions
Get the dimensions of the dataset (number of rows and columns).
```{r dimensions-data}
dim(wine.df)
```

## Initial Data Exploration
Check the column names of the dataset.
```{r colnames-data}
colnames(wine.df)
```

View the data structure
```{r str-data}
glimpse(wine.df)
```

Display the first 6 rows of the dataset to understand its structure and contents.
```{r head-data}
head(wine.df)
```

## Data Summary
Provide a summary of the dataset to get an overview of the variables and their distributions.
```{r summary-data}
summary(wine.df)
```

## Missing and Duplicated Values
Check for missing values in the dataset.
```{r missing-values}
colSums(is.na(wine.df))
```

Check for duplicated values in the dataset.
```{r duplicate-values}
dupes <- wine.df[duplicated(wine.df) | duplicated(wine.df, fromLast = TRUE), ]
# Print or inspect the duplicate entries
print(dupes)
```

## Data Cleaning
Clean the column names for easier handling.
```{r clean-data}
wine.df <- wine.df %>%
  janitor::clean_names()
```

## Data Skimming
Get a detailed summary of the dataset using the `skimr` package.
```{r skim-data}
skim(wine.df)
```
## Target Variable Distribution
Visualize the distribution of the target variable `quality`.
```{r target-distribution}
wine.df %>%
  ggplot(aes(x = quality, fill = factor(quality))) +
  geom_bar() +
  labs(title = "Distribution of Wine Quality Scores",
       x = "Quality Score",
       y = "Count",
       fill = "quality")
```

## Correlation Heatmap
Visualize the correlation matrix of numeric features using a heatmap.
```{r correlation-heatmap}
wine.df |>
  select(where(is.numeric)) |>
  cor() |>
  as.data.frame() |>
  rownames_to_column("feature") |>
  pivot_longer(-feature, names_to = "feature2", values_to = "corr") |>
  ggplot(aes(feature, feature2, fill = corr)) +
  geom_tile() +
  labs(title = "Correlation Heatmap (Numeric Features)", 
       x = "", 
       y = "") +
  scale_fill_viridis_c() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Check Business Insight Features
Analyze specific features for business insights.
```{r business-insights}
wine.df |>
  group_by(quality) |>
  summarise(
    n = n(),
    alcohol_mean = mean(alcohol, na.rm = TRUE),
    volatile_acidity_mean = mean(volatile_acidity, na.rm = TRUE),
    sulphates_mean = mean(sulphates, na.rm = TRUE),
    .groups = "drop"
    ) |>
  arrange(quality)
```


\newpage

# **Data Preparation**
## Data Splitting
Split the dataset into training and testing sets.
```{r data-splitting}
set.seed(123)
data_split <- initial_split(wine.df, prop = 0.8, strata = quality)
train_val <- training(data_split)
test_df <- testing(data_split)
```

Further split the training set into training and validation sets.
```{r}
data_split2 <- initial_split(train_val, prop = 0.8, strata = quality)
train_df <- training(data_split2)
val_df <- testing(data_split2)
```

## Display Data Dimensions
```{r data-dimensions}
kable(dimensions_data <- tibble(
  Dataset = c("Full Dataset", "Train + Validation Set", "Training Set", 
              "Validation Set", "Test Set"),
  Rows = c(nrow(wine.df), nrow(train_val), nrow(train_df), 
           nrow(val_df), nrow(test_df)),
  Columns = c(ncol(wine.df), ncol(train_val), ncol(train_df),
              ncol(val_df), ncol(test_df))
))
```

## Recipe Creation
Create a recipe for data preprocessing.
```{r data-preprocessing}
rec <- recipe(quality ~ ., data = train_df) |>
step_zv(all_predictors()) |>
step_normalize(all_predictors())
prep_rec <- prep(rec, training = train_df, retain = TRUE)
train_processed <- bake(prep_rec, new_data = train_df)
val_processed <- bake(prep_rec, new_data = val_df)
test_processed <- bake(prep_rec, new_data = test_df)
glimpse(train_processed)
```

## Convert to Matrix Format
Convert the processed data into matrix format for modeling.
```{r data-matrix}
x_train <- train_processed |> select(-quality) |> as.matrix()
y_train <- train_processed$quality
x_val <- val_processed |> select(-quality) |> as.matrix()
y_val <- val_processed$quality
x_test <- test_processed |> select(-quality) |> as.matrix()
y_test <- test_processed$quality
```

## Sanity Check
Perform a sanity check on the processed data.
```{r sanity-check}
dim(x_train); length(y_train)
dim(x_val); length(y_val)
dim(x_test); length(y_test)
```

<!-- ```{r end-document, include=FALSE} -->
<!-- knitr::knit_exit() -->
<!-- ``` -->


\newpage

# **Modeling**
## Baseline Metrics (Naive)
Calculate baseline metrics using the mean quality score from the training set.
```{r baseline-metrics}
baseline_pred <- rep(mean(y_train), length(y_test))
baseline_tbl <- tibble(
  truth = y_test,
  estimate = baseline_pred
)

baseline_tbl %>%
  summarize(
    RMSE = rmse_vec(truth, estimate),
    MAE = mae_vec(truth, estimate)
  )
```

## Neural Network Model (Regression)
```{r neural-network-model}
# neuralnet expects a data.frame with numeric columns and a formula
train_nn <- train_processed |> as.data.frame()
val_nn   <- val_processed   |> as.data.frame()
test_nn  <- test_processed  |> as.data.frame()

set.seed(123)

# Fit a feed-forward neural network for regression
# Note: smaller hidden layers keep training stable and fast on this dataset.
nn_fit <- neuralnet::neuralnet(
  quality ~ .,
  data = train_nn,
  hidden = c(10, 5),
  linear.output = TRUE,
  lifesign = "minimal",
  stepmax = 1e6
)

#nn_fit
```

## Predictions and Performance
```{r nn-predict-evaluate}
# Predict on validation and test sets
val_pred  <- neuralnet::compute(nn_fit, val_nn  |> dplyr::select(-quality))$net.result[, 1]
test_pred <- neuralnet::compute(nn_fit, test_nn |> dplyr::select(-quality))$net.result[, 1]

val_metrics <- tibble(truth = val_nn$quality, estimate = val_pred) |>
  summarise(
    RMSE = yardstick::rmse_vec(truth, estimate),
    MAE  = yardstick::mae_vec(truth, estimate)
  )

test_metrics <- tibble(truth = test_nn$quality, estimate = test_pred) |>
  summarise(
    RMSE = yardstick::rmse_vec(truth, estimate),
    MAE  = yardstick::mae_vec(truth, estimate)
  )

list(Validation = val_metrics, Test = test_metrics)
```

## Diagnostic Plots
```{r nn-diagnostics}
# Actual vs Predicted (Test)
plot_tbl <- tibble(
  truth = test_nn$quality,
  estimate = test_pred,
  residual = test_nn$quality - test_pred
)

p1 <- ggplot(plot_tbl, aes(x = truth, y = estimate)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Neural Network: Actual vs Predicted (Test)",
    x = "Actual Quality",
    y = "Predicted Quality"
  )

p2 <- ggplot(plot_tbl, aes(x = estimate, y = residual)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Predicted (Test)",
    x = "Predicted Quality",
    y = "Residual (Actual - Predicted)"
  )

p1
p2
```
<!-- ```{r end-document, include=FALSE} -->
<!-- knitr::knit_exit() -->
<!-- ``` -->


\newpage

# **Evaluation and Conclusion**
## Model Training




## Conclusion

 
 
\newpage

# **References**
1. UCI Machine Learning Repository: Wine Quality Data Set. https://archive.ics.uci.edu/ml/datasets/wine+quality
2. Fritsch, S., & Guenther, F. (2019). *neuralnet: Training of Neural Networks*. R package.
3. Kuhn, M., & Johnson, K. (2019). *Feature Engineering and Selection: A Practical Approach for Predictive Models*. CRC Press.
4. Wickham, H., et al. (2019). *Welcome to the tidyverse*. Journal of Open Source Software, 4(43), 1686.
5. Allaire, J., et al. (2024). *R Markdown: The Definitive Guide*. Chapman and Hall/CRC.
